---
title: "Getting Started with plotR"
author: "plotR Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with plotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction to plotR

plotR is an R interface to Observable Plot, a powerful JavaScript library for exploratory data visualization. This vignette will get you started with creating beautiful, interactive visualizations.

## Installation

```{r}
# Install from GitHub
devtools::install_github("yourusername/plotR")

# Load the package
library(plotR)
```

## Basic Concepts

plotR follows the **grammar of graphics** principle, where visualizations are built by combining:

- **Data**: Your R data frame
- **Aesthetics**: Visual properties (x, y, color, size)
- **Marks**: Geometric objects (dots, lines, bars)
- **Interactivity**: Tooltips and hover effects

## Your First Plot

Let's create a simple scatter plot using the built-in `mtcars` dataset:

```{r}
library(plotR)

# Basic scatter plot
plotR_scatter(
  data = mtcars,
  x = "wt",           # Weight on x-axis
  y = "mpg",          # Miles per gallon on y-axis
  title = "Fuel Efficiency vs Weight",
  save_as = "my_first_plot.html"
)
```

This creates an interactive HTML file that you can open in any web browser.

## Adding Visual Encoding

Make your plots more informative by encoding additional variables:

```{r}
# Add color and size encoding
plotR_scatter(
  data = mtcars,
  x = "wt", 
  y = "mpg",
  fill = "cyl",       # Color by number of cylinders
  size = "hp",        # Size by horsepower
  title = "Motor Cars: Multi-dimensional Analysis",
  save_as = "enhanced_scatter.html"
)
```

## Interactive Tooltips

All plotR charts include interactive tooltips. Customize them with the `tooltip_template` parameter:

```{r}
plotR_scatter(
  data = mtcars,
  x = "wt", 
  y = "mpg",
  fill = "cyl",
  tooltip_template = "🚗 Weight: ${d.wt * 1000} lbs\\n⛽ MPG: ${d.mpg}\\n🔧 ${d.cyl} cylinders",
  save_as = "custom_tooltips.html"
)
```

### Tooltip Syntax

- Use `${d.column_name}` to reference data columns
- Use `\\n` for line breaks
- Include emojis and formatting for visual appeal
- JavaScript expressions like `${d.wt * 1000}` are supported

## Chart Types

plotR supports various chart types:

### Line Charts
```{r}
# Time series data
sales_data <- data.frame(
  year = 2015:2023,
  revenue = c(100, 120, 140, 135, 160, 180, 170, 200, 220)
)

plotR_line(
  data = sales_data,
  x = "year",
  y = "revenue",
  title = "Revenue Growth",
  save_as = "line_chart.html"
)
```

### Bar Charts
```{r}
# Aggregate data first
cyl_summary <- aggregate(mpg ~ cyl, data = mtcars, FUN = mean)

plotR_bar(
  data = cyl_summary,
  x = "cyl",
  y = "mpg",
  title = "Average MPG by Cylinders",
  save_as = "bar_chart.html"
)
```

## Data Preparation

plotR works best with tidy data. Use dplyr for data manipulation:

```{r}
library(dplyr)
library(plotR)

mtcars %>%
  mutate(
    efficiency = case_when(
      mpg > 25 ~ "High",
      mpg > 15 ~ "Medium", 
      TRUE ~ "Low"
    )
  ) %>%
  plotR_scatter(
    x = "wt",
    y = "mpg", 
    fill = "efficiency",
    title = "Fuel Efficiency Categories"
  )
```

## Output Options

plotR creates standalone HTML files that are:

- **Portable**: Work in any web browser
- **Interactive**: Include hover tooltips and responsive design
- **Shareable**: Easy to email or host online
- **Embeddable**: Can be included in R Markdown documents

## Next Steps

- Read the [Advanced Charts](advanced-charts.html) vignette for complex visualizations
- Learn about [Customization](customization.html) for styling options
- Explore [Interactivity](interactivity.html) for advanced tooltip techniques

## Common Patterns

### Faceting by Group
```{r}
# Create separate plots by transmission type
manual_cars <- mtcars[mtcars$am == 1, ]
auto_cars <- mtcars[mtcars$am == 0, ]

plotR_scatter(manual_cars, "wt", "mpg", title = "Manual Transmission")
plotR_scatter(auto_cars, "wt", "mpg", title = "Automatic Transmission")
```

### Combining with Statistical Functions
```{r}
# Add trend analysis
library(broom)

# Fit linear model and create prediction data
model <- lm(mpg ~ wt, data = mtcars)
pred_data <- data.frame(
  wt = seq(min(mtcars$wt), max(mtcars$wt), length.out = 100)
)
pred_data$mpg <- predict(model, pred_data)

# Plot original data
plotR_scatter(
  data = mtcars,
  x = "wt", 
  y = "mpg",
  title = "Fuel Efficiency with Trend"
)

# Plot trend line separately
plotR_line(
  data = pred_data,
  x = "wt",
  y = "mpg", 
  title = "Predicted Trend"
)
```

## Troubleshooting

**Q: My plot appears blank**
A: Check that your data contains the specified column names and has no missing values in key columns.

**Q: Tooltips don't appear**
A: Ensure you're viewing the HTML file in a web browser, not just the source code.

**Q: Colors don't show up**
A: Make sure the `fill` parameter references a valid column name in your data.

## Summary

plotR makes it easy to create interactive, professional-quality visualizations with just a few lines of R code. The key concepts to remember:

1. Start with clean, tidy data
2. Map variables to visual aesthetics (x, y, fill, size)
3. Choose the appropriate chart type
4. Customize tooltips for interactivity
5. Save as HTML for sharing and embedding

Happy plotting! 📊✨